---
title: "Course Notes"
author: "L. Odette"
format: html
page-layout: full
toc: true
editor: visual
code-fold: true
code-line-numbers: true
code-block-border-left: "#31BAE9"
theme: sandstone
fontsize: 85%
self-contained: true
date: 12/26/2025
---

## Model Overview

BLP is a model of individuals making a discrete choice from different alternatives.

-   Choices are made in `markets` denoted by $t \in \mathcal{T}$.
    -   these are time periods, geographic regions, period-region pairs, etc.
-   Each market has `individuals` with types denoted by $i \in \mathcal{I}_t$.
    -   different demographics and preferences.
-   Individuals are faced with `choices` denoted by $j \in \mathcal{J}_t$.
    -   products, hospitals, candidates, etc.
    -   [**outside option**]{.underline} $j = 0$: no purchase, no treatment, no vote, etc.

## Utility Maximization

$$
\max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} 
$$

-   Individuals choose an alternative to maximize their (indirect) utility $u_{ijt}$.
    -   we will specify a function for $u_{ijt}$ and use revealed preferences to estimate it.
-   It'll help to decompose utility into three parts.
    -   mean utility $\delta_{jt}$: Average preference across all individuals in the market.
    -   systematic heterogeneity $\mu_{ijt}$: Different preferences, e.g. due to different demographics.
    -   idiosyncratic heterogeneity $\varepsilon_{ijt}$: Superimposed noise that accommodates estimation.

$$
\max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} =\delta_{jt} + \mu_{ijt} + \varepsilon_{ijt}
$$

-   We'll parameterize $\delta_{jt}$ and $\mu_{ijt}$ and make a convenient assumption about $\varepsilon_{ijt}$.

## Aggregate Market Shares

$$
\max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} = \delta_{jt} + \mu_{ijt} + \varepsilon_{ijt} \quad\implies\quad s_{ijt} = \Pr_{\varepsilon_{it}}\left(u_{ijt} \geq u_{ikt} \text{ for all } k \in \mathcal{J}_t \cup \{0\}\right)
$$

-   Assume a convenient distribution for $\varepsilon_{ijt}$: i.i.d. type I [extreme value](https://www.itl.nist.gov/div898/handbook/eda/section3/eda366g.htm) (or [Gumbel](https://en.wikipedia.org/wiki/Gumbel_distribution)) distribution[^1].
    -   **Logit shocks** are convenient because they give multinomial logit choice probabilities $s_{ijt}$.

[^1]: The Gumbel distribution is used to model the distribution of the maximum (or the minimum) of a number of samples of various distributions. In the [latent variable](https://en.wikipedia.org/wiki/Latent_variable "Latent variable") formulation of the [multinomial logit](https://en.wikipedia.org/wiki/Multinomial_logit "Multinomial logit") model — common in [discrete choice](https://en.wikipedia.org/wiki/Discrete_choice "Discrete choice") theory — the errors of the latent variables follow a Gumbel distribution. This is useful because the difference of two Gumbel-distributed [random variables](https://en.wikipedia.org/wiki/Random_variable "Random variable") has a [logistic distribution](https://en.wikipedia.org/wiki/Logistic_distribution "Logistic distribution").

$$
\max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} = \delta_{jt} + \mu_{ijt} + \varepsilon_{ijt} \quad\implies\quad s_{ijt} = \frac{\exp(\delta_{jt} + \mu_{ijt})}{\sum_{k \in \mathcal{J}_t \cup \{0\}} \exp(\delta_{kt} + \mu_{ikt})}
$$

-   Want $\mu_{ijt}$ to be sufficiently flexible that this convenient assumption matters little.

    -   Possible to eliminate $\varepsilon_{ijt}$ but computation gets difficult.

-   Each type $i$ is a share $w_{it}$ of the population. Aggregating over them gives **market shares**. $$
            s_{jt} = \sum_{i \in \mathcal{I}_t} w_{it} \cdot s_{ijt}
    $$

-   We'll match these to observed quantities $q_{jt} = s_{jt} \cdot M_t$ in our data.

## Why Type I extreme value?

### Computational Tractability
The primary advantage is that Type I extreme value errors lead to closed-form choice probabilities through the logit formula. When consumer $i$'s utility from product $j$ in market $t$ is:
$$
u_{ijt} = δ_{jt} + μ_{ijt} + ε_{ijt}
$$
where $ε_{ijt}$ follows Type I extreme value, the choice probability integrates to the familiar logit form conditional on the random coefficients. This gives us:

$$
P_{ijt} = exp(\delta_{ijt} + \mu_{ijt}) / \sum_k exp(δ_{kt} + \mu_{ikt})
$$

Without this assumption, we'd need to evaluate multidimensional integrals for each choice probability, making estimation computationally infeasible for markets with many products.

## Type I Extreme Value Distribution Properties 

The Type I extreme value (Gumbel) distribution has the following characteristics:

**Cumulative Distribution Function (CDF):**

$$F(\varepsilon) = \exp(-\exp(-\varepsilon))$$

**Probability Density Function (PDF):**

$$f(\varepsilon) = \exp(-\varepsilon - \exp(-\varepsilon))$$

These specific functional forms are what enable the closed-form solution for choice probabilities.

### Derivation of Choice Probability

#### Choice Rule

Consumer $i$ chooses product $j$ if and only if it provides the highest utility among all available alternatives:

$$\text{Choose } j \iff u_{ij} > u_{ik} \quad \forall k \neq j$$

#### Conditional Choice Probability

Given the deterministic components $\{\delta, \mu_i\}$, the choice probability is:

$$P_{ij}|\mu_i = \Pr(\varepsilon_{ij} > \varepsilon_{ik} + \delta_k + \mu_{ik} - \delta_j - \mu_{ij}, \quad \forall k \neq j)$$

Let's define $V_{ij} = \delta_j + \mu_{ij}$ as the deterministic part of utility. This allows us to rewrite:

$$P_{ij}|\mu_i = \Pr(\varepsilon_{ik} < \varepsilon_{ij} + V_{ij} - V_{ik}, \quad \forall k \neq j)$$

#### Probability Conditional on $\varepsilon_{ij}$

Given a specific realization $\varepsilon_{ij} = \varepsilon$, the probability that product $j$ is chosen is the probability that all other alternatives have lower utility:

$$\Pr(j \text{ chosen}|\varepsilon_{ij} = \varepsilon) = \prod_{k \neq j} F(\varepsilon + V_{ij} - V_{ik})$$

Substituting the Type I extreme value CDF:

$$= \prod_{k \neq j} \exp(-\exp(-(\varepsilon + V_{ij} - V_{ik})))$$

Using properties of exponentials:

$$= \prod_{k \neq j} \exp(-\exp(-\varepsilon)\exp(V_{ik} - V_{ij}))$$

$$= \exp\left(-\exp(-\varepsilon)\sum_{k \neq j}\exp(V_{ik} - V_{ij})\right)$$

### Integration Over the Idiosyncratic Shock

#### Setting Up the Integral

To obtain the unconditional (on $\varepsilon_{ij}$) choice probability, we integrate over the distribution of $\varepsilon_{ij}$:

$$P_{ij}|\mu_i = \int_{-\infty}^{\infty} \exp\left(-\exp(-\varepsilon)\sum_{k \neq j}\exp(V_{ik} - V_{ij})\right) \cdot \exp(-\varepsilon - \exp(-\varepsilon)) \, d\varepsilon$$

#### Change of Variables

We employ the substitution $t = \exp(-\varepsilon)$, which gives us:

- $dt = -\exp(-\varepsilon)d\varepsilon = -t \cdot d\varepsilon$
- Therefore: $d\varepsilon = -\frac{dt}{t}$
- When $\varepsilon \to -\infty$: $t \to \infty$
- When $\varepsilon \to \infty$: $t \to 0$

#### Transformed Integral

The integral becomes:

$$P_{ij}|\mu_i = \int_{\infty}^{0} \exp\left(-t\sum_{k \neq j}\exp(V_{ik} - V_{ij})\right) \cdot \exp(-t) \cdot (-dt)$$

Reversing the limits of integration:

$$= \int_{0}^{\infty} \exp\left(-t\left[1 + \sum_{k \neq j}\exp(V_{ik} - V_{ij})\right]\right) \, dt$$

#### Simplification

Let $A = 1 + \sum_{k \neq j}\exp(V_{ik} - V_{ij}) = \sum_{k=1}^{J}\exp(V_{ik} - V_{ij})$ (where the sum now includes all products including $j$)

Then:

$$P_{ij}|\mu_i = \int_{0}^{\infty} \exp(-tA) \, dt = \left[-\frac{1}{A}\exp(-tA)\right]_{0}^{\infty} = \frac{1}{A}$$

$$= \frac{1}{\sum_{k}\exp(V_{ik} - V_{ij})}$$

### The Final Logit Form

#### Conditional Logit Probability

Multiplying both numerator and denominator by $\exp(V_{ij})$:

$$P_{ij}|\mu_i = \frac{\exp(V_{ij})}{\sum_{k}\exp(V_{ik})} = \frac{\exp(\delta_j + \mu_{ij})}{\sum_{k}\exp(\delta_k + \mu_{ik})}$$

This is the **multinomial logit formula conditional on the random coefficients** $\mu_i$.

#### Unconditional Choice Probability in BLP

The unconditional choice probability requires one additional integration over the distribution of random coefficients:

$$P_{ij} = \int \frac{\exp(\delta_j + \mu_{ij})}{\sum_{k}\exp(\delta_k + \mu_{ik})} \, dF_{\mu}(\mu_i)$$

This final integral typically lacks a closed-form solution and requires simulation methods (e.g., Monte Carlo integration), but the Type I extreme value assumption has already provided the crucial computational advantage by yielding the tractable logit kernel.



### Preservation of Heterogeneity
Unlike the simple logit model, BLP doesn't rely solely on Type I extreme value shocks for consumer heterogeneity. The model includes:

Random coefficients ($\mu_{ijt}$) that capture observed and unobserved heterogeneity in preferences
Type I extreme value shocks ($\varepsilon_{ijt}$) that capture remaining idiosyncratic variation

This two-layer approach allows the model to overcome the restrictive substitution patterns of simple logit (IIA property) while maintaining computational feasibility.

### Consistency with Random Utility Theory

The Type I extreme value distribution has deep roots in discrete choice theory dating back to McFadden (1974). It emerges naturally from several micro-foundations:

It can be derived as the limiting distribution of the maximum of many random variables
It's consistent with a random utility maximization framework
It satisfies the independence from irrelevant alternatives property conditional on the random coefficients

### Smooth Aggregation
The smooth functional form of the logit probabilities ensures that market shares are continuously differentiable with respect to parameters. This is essential for:

Applying the BLP inversion (contraction mapping) to recover mean utilities δ_j
Computing gradients for optimization algorithms
Ensuring the model is well-behaved econometrically

### Empirical Flexibility
While the Type I extreme value assumption might seem restrictive, in practice:

The random coefficients μ_ij handle most of the important substitution patterns
The ε_ij shocks mainly serve to smooth out the choice probabilities and ensure positive market shares for all products
The variance of these shocks (typically normalized to π²/6) provides a natural scale for utility

## Choosing Market Size

-   In our data, we observe quantities $q_{jt} = s_{jt} \cdot M_t$.
    -   Need to divide by some market size $M_t$ to get our model's market shares $s_{jt}$.
    -   Issue here is that we often don't observe the quantity of outside choices $q_{0t}$.
-   Sometimes the choice of market size ([potential]{.underline} demand) is straightforward.
    -   Market size for drugs to treat a condition is how many people have that condition.
-   But typically, the choice of market size is neither **clear nor innocuous**.
    -   E.g. how many choices of which cereal to buy are made every day in a specific city?
    -   Population $\times$ max cereals per day? Foot traffic estimate $\times$ max cereals per trip?

## Identification and Normalizations

$$
 \max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} = \delta_{jt} + \mu_{ijt} + \varepsilon_{ijt}
$$ -

-   We'll estimate our utility function with **revealed preferences**. -

    -   Holding $\mu_{ijt}$ fixed, a higher quantity $q_{jt} > q_{kt}$ implies a higher mean utility $\delta_{jt} > \delta_{kt}$.

-   Utility is invariant to positive affine transformations. Need two normalizations. $$u_{ijt} > u_{ikt} \quad \xLeftrightarrow{b > 0}\quad a + b \cdot u_{ijt} > a + b \cdot u_{ikt}$$

    a.  **Level:** We will normalize $u_{i0t} = \varepsilon_{i0t}$, i.e. $\delta_{0t} = \mu_{i0t} = 0$ .$\implies$ Estimates are relative to outside option utility.
    b.  **Scale:** We already normalized $\text{Var}(\varepsilon_{ijt}) = \pi^2 / 6$ when deriving choice probabilities. $\implies$ Estimates are relative to scale of noise.

    -   Now that our model can in theory be identified, how do we estimate it?

## Pure Logit Model

$$
\max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} = \delta_{jt} + \cancelto{0}{\mu_{ijt}} + \varepsilon_{ijt} 
$$

-   Start with the simplest case: no heterogenous utility.
-   Market shares simplify. No aggregation over individual types.

$$
\max_{j \in \mathcal{J}_t \cup \{0\}} u_{ijt} = \delta_{jt} +\varepsilon_{ijt}\quad\implies\quad s_{jt}=\frac{\exp \delta_{jt}}{1+\sum_{k \in \mathcal{J}_t} \exp \delta_{kt}}\quad\implies\quad \log\frac{s_{jt}}{s_{0t}}=\delta_{jt}
$$

-   We can recover mean utilities from observed market shares \citep{berry1994estimating}.
    -   If we specify a function for $\delta_{jt}$, we'll have a linear regression!

## Pure Logit Estimating Equation

$$
\log\frac{s_{jt}}{s_{0t}}=\delta_{jt}= \alpha p_{jt} + x_{jt}'\beta+\xi_{jt}
$$ \## Interpreting Parameters

-   Let's say we estimate this equation. How to interpret our parameter estimates?
-   Prices are in dollars, so the units of $\alpha$ are \`\`utils'' per dollar. Not very helpful.
    -   Instead, report own-price elasticities, or a quantity-weighted average/median.
    -   You can derive elasticities by differentiating the multinomial logit expression for $s_{jt}$.

$$
   \eta_{jjt} = \frac{\partial\log q_{jt}}{\partial\log p_{jt}} = \frac{\partial q_{jt}}{\partial p_{jt}} \frac{p_{jt}}{q_{jt}} = \frac{\partial s_{jt}}{\partial p_{jt}} \frac{p_{jt}}{s_{jt}} = \alpha \cdot p_{jt} \cdot (1 - s_{jt})
$$

-   If $x_{jt}$ is a `mushy` cereal dummy, $\beta$ is `utils` from mushyness. Again, not helpful.
    -   Instead, report $\beta / \alpha$, the dollar willingness to pay for mushyness.

## Exercise 1

```{r}
#| echo: false
#| message: false
#| label: initialization
require(ggplot2)
require(patchwork)

source('../R/conda_init.R')

theme_set(theme_bw(base_size = 18) + theme(legend.position = "top"))

# https://medium.com/save-the-data/how-to-use-python-in-r-with-reticulate-and-conda-36685534f06a
require(reticulate)
Sys.setenv(RETICULATE_PYTHON="../envs/bin/python")
```

```{python}
#| label: import python packages
#| code-summary: import python packages
import pyblp
import numpy as np
import pandas as pd
import statsmodels.formula.api as smf
```

```{r}
#| label: read product data
#| code-summary: read product data
dat <- 
  readr::read_csv("../Exercises/Data/products.csv", show_col_types = FALSE)
```

| Column | Data Type | Description |
|------------------------|------------------------|------------------------|
| `market` | String | The city-quarter pair that defines markets $t$ used in these exercises. The data were motivated by real cereal purchase data across 47 US cities in the first 2 quarters of 1988. |
| `product` | String | The firm-brand pair that defines products $j$ used in these exercises. Each of 5 firms produces between 1 and 9 brands of cereal. |
| `mushy` | Binary | A dummy product characteristic equal to one if the product gets soggy in milk. |
| `servings_sold` | Float | Total quantity $q_{jt}$ of servings of the product sold in a market, which will be used to compute market shares. |
| `city_population` | Float | Total population of the city, which will be used to define a market size. |
| `price_per_serving` | Float | The product's price $p_{jt}$ used in these exercises. |
| `price_instrument` | Float | An instrument to handle price endogeneity in these exercises. Think of it as a cost-shifter, a Hausman instrument, or any other valid IV that we discussed in class. |

```{r}
#| label: describe data
#| code-summary: describe data
dat |> summary()
```

```{r}
#| label: create shares
#| code-summary: create shares

share_dat <- dat |> 
  dplyr::mutate(
    market_size = city_population * 90            # estimate market size
    , market_share = servings_sold / market_size  # compute share quantity sold
  ) |> 
  dplyr::group_by(market) |>                      # group by market
  dplyr::mutate(
    inside_share_sum = sum(market_share)          # sum inside shares
    , outside_share = 1 - inside_share_sum        # calculate outside share (i.e. none of above)
  ) |> 
  dplyr::ungroup()

share_dat |>                                      # summarize shares across markets
  dplyr::summarize(
    min_s0t = min(outside_share)
    , max_s0t = max(outside_share)
  )

```

```{r}
#| label: compute logit share and then run regression
#| code-summary: compute logit share and then run regression
#
model_dat <- share_dat |> 
  dplyr::mutate(
    logit_delta = log(market_share / outside_share)
  )

lm(logit_delta ~ 1 + price_per_serving + mushy, data = model_dat) |> broom::tidy()
```

```{r}
#| label: rename columns for use in pyblp
#| code-summary: rename columns for use in pyblp
py_model_dat <- 
  model_dat |> dplyr::rename(
    market_ids = market
    , product_ids = product
    , shares = market_share
    , prices = price_per_serving
  ) |> 
  dplyr::mutate(demand_instruments0 = prices)
```

```{python}
#| label: formulate problem for pyblp
#| code-summary: formulate problem for pyblp
ols_problem = pyblp.Problem(pyblp.Formulation('1 + mushy + prices'), r.py_model_dat)
```

```{python}
#| label: not sure
#| code-summary: not sure
ols_problem.products.X1

ols_problem.products.ZD
```

```{python}
#| label: solve problem with pyblp
#| code-summary: solve problem with pyblp
ols_results = ols_problem.solve(method='1s')
```

```{r}
#| label: check instrument
#| code-summary: check instrument
#
lm(prices ~ price_instrument + factor(market_ids) + factor(product_ids), data = py_model_dat) |> broom::tidy()

```

```{r}
#| label: rename columns again for use in pyblp with instrument
#| code-summary: rename columns again for use in pyblp with instrument
#
py_model_dat_1 <- 
  model_dat |> dplyr::rename(
    market_ids = market
    , product_ids = product
    , shares = market_share
    , prices = price_per_serving
  ) |> 
  dplyr::mutate(demand_instruments0 = price_instrument)
```

```{python}
#| label: formulate problem with fixed effects for pyblp
#| code-summary: formulate problem with fixed effects for pyblp
#
ols_problem = pyblp.Problem(
  pyblp.Formulation('prices', absorb='C(market_ids) + C(product_ids)'), r.py_model_dat_1)
```

```{python}
#| label: solve problem with fixed effects for pyblp
#| code-summary: solve problem with fixed effects for pyblp

ols_results = ols_problem.solve(method='1s')
```

```{r}
#| label: generate counterfacual data
#| code-summary: generate counterfacual data

counterfactual_data <- py_model_dat_1 |> 
  dplyr::filter(market_ids == "C01Q2")
counterfactual_data
```

```{r}
#| label: filter counterfacual data
#| code-summary: filter counterfacual data
counterfactual_data |> 
  dplyr::filter(stringr::str_starts(product_ids,"F1")) |> 
  dplyr::pull(product_ids) |> unique()
```
